---
title: "General Social Survey"
author: "Matias Lucha"
date: "June 2024"
output: 
  html_document:
    number_sections: true
    toc: true
---


```{r include=FALSE,echo=FALSE}
require(tigerstats)
require(tidyverse)
require(foreign)
source("GrabData.R")
```




#  Question 1

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

The data in the dataframe GSSdata is from the 2018 General Social Survey. The first blocks of R-code has selected down a subset of the data to just 16 variables. It has further removed unwanted factor levels in much of the data. Examine the code in the GrabData.R file to see what it is doing. Some of the variables are categorical and others are numerical. Be sure to do a variable analysis before tackling each question.  
First question - Is opinion on the death penalty (CAPPUN) independent of gun ownership (OWNGUN)?

$H_0$: Opinion on capital punishment is independent of gun ownership.

$H_A$: Capital punishment and gun ownership are dependent. 

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

OWGUN is categorical, CAPPUN is also categorical.

We want CAPPUN ~ OWNGUN
The helper file we use is CAT~CAT

## Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
barchartGC(~CAPPUN + OWNGUN,data=GSSdata)
barchartGC(~CAPPUN + OWNGUN,data=GSSdata, type="percent")

```

This graph shows that among people who own a gun, are more likely to favor capital punishment.

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
table1 <- xtabs(~CAPPUN + OWNGUN, data=GSSdata)
rowPerc(table1)
colPerc(table1)
```
Numerical summary shows that 75% of gun owners favor CAPPUN.

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```
The Chi-square test p-value is 0.0202, Hence we reject the null hypothesis and conclude that gun ownership and capital punishment are not independent. 

The fisherexact test has a p-value of 0.01651, Hence we reject the null hypothesis, as with the chi-square test, and furthermore the odds ratio says that gunowners are 2.3 times more likely to favor capital punishment than those who don't own a gun. 


#  Question 2

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Is there a difference in years of education (EDUC) between the those who did or did not vote in 2012 (VOTE12)?

$H_0$: There is a difference in years of education between those who voted and did not vote in 2012.

$H_A$: There is no difference betwen education and if someone voted.

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

EDUC is categorical, VOTE12 is also categorical.

We want EDUC ~ VOTE12
The helper file we use is CAT~CAT

## Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
barchartGC(~EDUC + VOTE12,data=GSSdata)
barchartGC(~EDUC + VOTE12,data=GSSdata, type="percent")

```
This shows that there was an increase of people who voted in 2012 tended to have higher education.

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
table1 <- xtabs(~EDUC + VOTE12, data=GSSdata)
rowPerc(table1)
colPerc(table1)
```
The more education you have, the more likely you are to VOTE12. 
### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
t.test(EDUC~VOTE12, data=GSSdata)
```

This T TESTS shows that this means that those who voted had a better education. Therefore we reject the null hypothesis. 

The p-value is 0.000001248 which is less than 0.05, therefore we reject the null hypothesis. 

#  Question 3

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Determine if a persons performance on the wordsum test (WORDSUM) is independent of their level of schooling (DEGREE).

$H_0$: Performance on the wordsum test is independent of their level of schooling.

$H_A$: Performance on the wordsum test is dependent of their level of schooling.

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

WORDSUM is a numerical variable,
DEGREE us a categorical variable,
so I use the num~cat data file

## Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

```{r}
boxplot(WORDSUM~DEGREE,data=GSSdata)
```
This chart shows the level of schooling has an affect on the WORDSUM test percentage results.

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
barchartGC(~WORDSUM + DEGREE,data=GSSdata)
barchartGC(~WORDSUM + DEGREE,data=GSSdata, type="percent")

```

The charts show that level of schooling has an affect on the WORDSUM test percentage results. 

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
favstats(WORDSUM~DEGREE,data=GSSdata)
```
The data shows that performance on the wordsum test is dependent on the level of schooling, as the values get higher overall with education. 

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
anova=aov(WORDSUM~DEGREE, data=GSSdata)
summary(anova)
```
As we can see, the p-value is 0.00000025 which is smaller than 0.05. Therefore, we reject the null hypothesis.

```{r}
#we specify the margins starting at the bottom and proceeding clockwise - thus 8.7 is the left side margin to hold the names of the factors
par(mar = c(7, 8.7, 2, 2))
plot(TukeyHSD(anova),las=2)
```
The family-wise confidence levels for high school and less than high school have no difference. There is a difference between bachelor's degree and high school. There is a difference between graduate and high school. There is no difference between highschool and gradyate. There is no difference between jr college and graduate. There is no difference between bachelors and graduates. 



# Discussion and Conclusion

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->

#### Question 1: is gun ownership independent of opinion on capital punishment? 
The summary results and the inferential results show that they are not independent.
